WED-SQL (Draft)

### 2016-05-17 doc/paper/main_en.tex
paper: ch2

Most of modern business process management software systems are subject to recurrent structural modifications during its
life cycle. This changes are necessary to incorporate new requisites to the software, once it is often impossible 
to come up with all requirements upfront in the modeling phase of project - some of then only appear after the software
is running in production. Furthermore, it is too expensive and time consuming to try to predict all the possibles situations 
and its outcomes, which ultimately may raise exceptions that must be dealt with. Over time, these modifications leads
to code deterioration, exponentially increasing the maintenance costs and compromising the quality of the system as well 
negatively impacting its overall performance.
\\
\indent  Classic business process specification models, for example WS-BPEL, Process Algebra and Petri Networks, focus primarily
on behavioral interactions and relations between process, pushing data analysis into the background and overlooking its
relevance for the project. Therefore, none of these models exactly fits the needs of design dynamic business process 
management systems, which may result in costly system adaptations and maintenance.
\\
\indent  The WED-flow approach, in contrast with the classical models, capture both the inter-process relations and, with the very
same relevance, the data generated by their interactions. With this in mind, the WED-flow allows systems architects to 
think in terms of data states and transitions, similarly to a finite automata designing. Modifying or adding new business 
rules to the project, which is often translated in creation of new processes, is a simple matter of defining new data states 
and conditions for transitions, thus capturing the true nature of dynamic processes and simplifying an incremental evolution 
of the whole project. Real time exceptions are also easier to be handled and integrated to the project, once this approach
is based on transactional properties, thus providing several recovery mechanisms~\cite{ICWS12}.
\\
%problemas da implementacao e justificativa do WED-SQL
%tolerancia a falhas
\indent In order to provide a standard implementation as well as simplifying its software development, in this work we present 
the WED-SQL, an WED-flow framework. The main purpose of WED-SQL is to implement all WED-flow definitions~\cite{FTPM10} and encapsulate 
the control structure, what includes the transactional control, exceptions handling, data states detection, state transitioning 
and ensure data integrity. This framework effectively is an abstraction layer that allow developers to put more effort
on business logic and come up with better WED-flow projects, also making better use of reusable code. 
\\
\indent The WED-SQL framework has been built inside the PostgreSQL relational database management system, thereby uses SQL language
to specify WED-flow elements, has full transactional support and it is highly fault tolerant. Once it employs well consolidate 
technologies, widespread in the computer science field, this framework targets to disseminate the adoption of WED-flow on
business process management software design, by means of an easy to use, reliable and scalable tool that is capable to 
provide the flexibility demanded by modern transaction control systems. 
\\
\indent The remaining of this paper is structured as follows: Section II presents WED-SQL general software architecture aspects and
internal data structures. Section III briefly describes how the system works. A discussion about the algorithms involved
is presented in section IV. Following, a detailed description of transactional control and the communication protocol used
for external communication is given in section V. Finally, on section VI, we discuss the next challenges to improve and
expand this framework capabilities.

@@

the WED-trigger $g$ that, in turn, will initiate a WED-transition $t$ generating the next WED-state. This keeps going 
until a final WED-state is reached, when this instance is then finalized. Roughly speaking, a WED-flow application is a 
transactional data state transitioning machine.

@@

\subsection{WED-worker e WED-server}
O WED-server é, basicamente, uma extensão para o SGBD (Sistema Gerenciador de Banco de Dados) PostgreSQL~\cite{PSQL} composta de 
\emph{triggers}, tabelas de controle e \emph{stored procedures}~\cite{NAV}. A escolha do Postgresql como base do WED-server 
se deu por diversos motivos, dentre eles: ser de código aberto, implementação modular baseada em catálogs (\emph{system catalog driven}), 
facilidade de estender suas funcionalidades por meio de módulos escritos nas linguagens C, Python e SQL.

@@

\indent Por ser de código aberto, a licença do Postgresql garante que o mesmo possa ser modificado e redistribuído, o que permite
que o WED-SQL possa ser distribuído sem grandes empecilhos. Além disso, o acesso ao código fonte possibilita uma melhor compreensão
de seus mecanismos internos, garantido uma implementação mais robusta e eficiente do paradigma WED-flow. 

@@

\indent O PostgreSQL armazena seus dados de controle em tabelas especiais denominadas catálogos de sistema. Esses catálogos, 
que são facilmente acessados com as devidas permissões, expõe aos usuários informações vitais para estender suas funcionalidades
por meio da criação de módulos que podem ser carregados tanto de modo dinâmico, ou seja, com o SGBD executando ou de modo
estático, durante a inicialização. Dados de transações podem ser facilmente obtidos por meio desses 
catálogos de sistema, o que é particularmente útil no caso do WED-server. 

@@

\indent Devido à natureza dinâmica do paradigma WED-flow, a utilização de uma linguagem de programação de alto nível, como Python, 
é fundamental para que toda a sua expressividade seja implementada de forma plena. Do ponto de vista de implementação, 
o WED-server é composto de módulos escritos em Python, SQL e C, além de aproveitar o arcabouço transacional clássico (propriedades ACID, controle de concorrência e etc) fornecido pelo PostgreSQL.

>>>>

Most of modern business process management software systems are subject to recurrent structural modifications during its
life cycle. This changes are necessary to incorporate new requisites to the software, once it is often impossible 
to come up with all requirements upfront in the modeling phase of project - some of then only appear after the software
is running in production. Furthermore, it is too expensive and time consuming to try to predict all the possibles situations 
and its outcomes, which ultimately may raise exceptions that must be dealt with. Over time, these modifications leads
to code deterioration, exponentially increasing the maintenance costs and compromising the quality of the system as well 
negatively impacting its overall performance.
\\
\indent  Classic business process specification models, for example WS-BPEL, Process Algebra and Petri Networks, focus primarily
on behavioral interactions and relations between process, pushing data analysis into the background and overlooking its
relevance for the project. Therefore, none of these models exactly fits the needs of design dynamic business process 
management systems, which may result in costly system adaptations and maintenance.
\\
\indent  The WED-flow approach, in contrast with the classical models, capture both the inter-process relations and, with the very
same relevance, the data generated by their interactions. With this in mind, the WED-flow allows systems architects to 
think in terms of data states and transitions, similarly to a finite automata designing. Modifying or adding new business 
rules to the project, which is often translated in creation of new processes, is a simple matter of defining new data states 
and conditions for transitions, thus capturing the true nature of dynamic processes and simplifying an incremental evolution 
of the whole project. Real time exceptions are also easier to be handled and integrated to the project, once this approach
is based on transactional properties, thus providing several recovery mechanisms~\cite{ICWS12}.
\\
%problemas da implementacao e justificativa do WED-SQL
%tolerancia a falhas
\indent In order to provide a standard implementation as well as simplifying its software development, in this work we present 
the WED-SQL, an WED-flow framework. The main purpose of WED-SQL is to implement all WED-flow definitions~\cite{FTPM10} and encapsulate 
the control structure, what includes the transactional control, exceptions handling, data states detection, state transitioning 
and ensure data integrity. This framework effectively is an abstraction layer that allow developers to put more effort
on business logic and come up with better WED-flow projects, also making better use of reusable code. 
\\
\indent The WED-SQL framework has been built inside the PostgreSQL~\cite{PSQL} relational database management system, thereby uses SQL language
to specify WED-flow elements, has full transactional support and it is highly fault tolerant. Once it employs well consolidate 
technologies, widespread in the computer science field, this framework targets to disseminate the adoption of WED-flow on
business process management software design, by means of an easy to use, reliable and scalable tool that is capable to 
provide the flexibility demanded by modern transaction control systems. 
\\
\indent The remaining of this paper is structured as follows: Section II presents WED-SQL general software architecture aspects and
internal data structures. Section III briefly describes how the system works. A discussion about the algorithms involved
is presented in section IV. Following, a detailed description of transactional control and the communication protocol used
for external communication is given in section V. Finally, on section VI, we discuss the next challenges to improve and
expand this framework capabilities.

@@

the WED-trigger $g$ that, in turn, will initiate a WED-transition $t$, which must complete by some predefined timeout,thus 
generating the next WED-state. This keeps going until a final WED-state is reached, when this instance is then finalized. 
Roughly speaking, a WED-flow application is a transactional data state transitioning machine.

@@

\subsection{WED-worker and WED-server}
Fundamentally, the WED-server is a PostgreSQL extension package consisting of \emph{database triggers}, control tables and 
stored procedures~\cite{NAV}. Choosing PostgreSQL as the WED-server foundation was a decision based on several of its
aspects, such as be an open source project, catalog-driven operation, support for user-written code using low and high level 
programming languages (e.g., C, Python, Perl, TCL and SQL itself).

@@

\indent  WED-SQL directly benefits from the fact that, being an Open-source software, PostgreSQL's license states that it can be used, 
modified, copied and distributed without any kind of payment or agreement. Besides that, having access to the source code
makes development less complex and error prone once it is easier to analyze and, consequently, better comprehend its internal
mechanisms. 

@@

\indent Relational database systems needs to store metadata about internal structure of tables, columns, indexes, etc. This data
is kept on what is know as system catalogs. PostgreSQL not only stores much of its control data in these catalogs
but also base its procedures on then, in what is called a catalog-driven operation. Since they are readily  
available to users as ordinary tables, thus can be modified, it is fairly easy to modify the PostgreSQL behavior. This
property is specially useful to allow the WED-server to manage timed transactions related to WED-transitions and exception 
handling. 

@@

\indent As previously mentioned, users can write PostgreSQL modules using a few different languages. The most simple and straight
forward of then is a SQL dialect called pgSQL, that is basically SQL with variables and control structures. Also, modules
can be written in C language and loaded as a dynamic library, providing the user with ultimate logic control and performance.
Despite these two language having native support in PostgreSQL, sometimes pgSQL is not expressive enough when the logic
involved is too complex, while C, albeit fast, is too low level and overloaded with technical implementation details too
perform most of the tasks. The very PostgreSQL documentation recommends using a language other than C unless it is explicit
needed. Fortunately, it also offers support for Python, a much more versatile and expressive programming language than
the two former. Given the dynamic nature of WED-flow models, Python has been proven the perfect tool to do most of the job
on the WED-server side, only recurring to C when critical or where an extra performance is absolutely necessary. And,of
course, SQL statements are every where in both WED-server and WED-workers. 
